Computing Entropies with Nested Sampling

### 2017-10-17 talk/ms.tex
lots of writing

@@

>>>>

% New slide
\begin{frame}
\frametitle{Interpretation of conditional entropies}

How uncertain would the question ``what's the value of $\theta$,
precisely?'' be if the question ``what's the value of $d$, precisely''
were to be resolved?

@@

% New slide
\begin{frame}
\frametitle{Hard integrals}

E.g.
\begin{align}
H(\theta|d) &= -\int p(d) \int p(\theta|d) \log p(\theta|d) \, d\theta \, dd \\
            &= -\int p(d) \int p(\theta|d)
                    \log\left[\frac{p(\theta)p(d|\theta)}{p(d)}\right] \, d\theta \, dd
\end{align}

But $p(d)$, sitting there inside a logarithm, is already supposed to be
a hard integral (the marginal likelihood / evidence)...

\begin{align}
p(d) &= \int p(\theta) p(d|\theta) \, d\theta
\end{align}

% New slide
\begin{frame}
\frametitle{Marginal Likelihood Integral}
Nested Sampling was invented in order to do this hard integral
\begin{align}
p(d) &= \int p(\theta) p(d|\theta) \, d\theta
\end{align}

or
\begin{align}
Z &= \int \pi(\theta) L(\theta) \, d\theta
\end{align}
where $\pi=$ prior, $L=$ likelihood.
It's just an expectation. Why is it hard?
\end{frame}

% New slide
\begin{frame}
\frametitle{Simple Monte Carlo fails}

\begin{align}
Z &= \int \pi(\theta) L(\theta) \, d\theta \\
  &\approx \frac{1}{N} \sum_{i=1}^N L(\theta_i)
\end{align}
with $\theta_i \sim \pi$ will probably miss the
tiny regions where $L$ is high.

Equivalently, $\pi$ implies a {\em very heavy-tailed}
distribution of $L$-values, and simple Monte Carlo fails.

% New slide
\begin{frame}
\frametitle{Nested Sampling}
Nested Sampling takes the original problem and constructs
a 1D problem from it.

\begin{align}
Z = \int_0^1 L(X) dX
\end{align}
where
\begin{align}
X(\ell) &= \int_{L(\theta) > \ell} \pi(\theta) \, d\theta
\end{align}

\begin{block}{The meaning of $X$}
$X(\ell)$ is the amount of prior mass whose likelihood exceeds
$\ell$. As $\ell$ increases, $X$ decreases.
\end{block}

\begin{frame}
\frametitle{Nested Sampling}

\begin{center}
\includegraphics[width=0.8\textwidth]{skilling.png}

Figure from Skilling (2006).
\end{center}

Since $X(\ell)$ is the CDF of $L$-values implied by $\pi$,
points $\theta \sim \pi$ have a uniform distribution over $X$.

% New slide
\begin{frame}
\frametitle{Nested Sampling}
The idea is to generate a sequence of points with increasing
likelihoods, such that we can estimate their $X$ values. Since
we know their $L$-values, we can then do the integral 
numerically.

\begin{align}
Z = \int_0^1 L(X) dX
\end{align}

% New slide
\begin{frame}
\frametitle{Nested Sampling algorithm}

\begin{enumerate}
  \item<2-> Generate $N$ points from $\pi$
  \item<3-> Find the worst one (lowest likelihood $L^*$), save it
  \item<4-> Estimate its $X$ using Beta$(N,1)$ distribution\footnote{From order statistics.}
  \item<5-> Record worst particle and its $X$-value, then discard it.
  \item<6-> Replace that point with a new one from $\pi$ but with
        likelihood above $L^*$.
  \item<7-> Repeat steps 2--5 indefinitely.
\end{enumerate}
